---
title: "Class12"
author: "Joe Kesler"
format: html
---
# 2.Import count and col data
```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
head(counts)
head(metadata)
```
> Q1. How many genes are in this dataset?

```{r}
nrow(counts)
```

There are 38,694 genes in the dataset!

> Q2. How many ‘control’ cell lines do we have?

```{r}
dim(metadata)
```

There are 4 control cell lines.

# 3. Toy differential gene expression

Mean counts per gene across these samples:

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

Another way of calculating it:

```{r}
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```
> Q3. How would you make the above code in either approach more robust?

I could turn it into a function. 

> Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r}
library(dplyr)
treated <- metadata %>% filter(dex=="treated")
treated.counts <- counts %>% select(treated$id) 
treated.mean <- rowSums(treated.counts)/4
head(treated.mean)
```

We will combine our meancount data for bookeeping purposes

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

> Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

```{r}
plot(meancounts)
```

> Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. 

```{r}
library(ggplot2)
ggplot(meancounts, aes(x = control.mean, y = treated.mean, alpha = 0.2)) +
  geom_point()
```
> Q6. Try plotting both axes on a log scale. 

```{r}
library(ggplot2)
ggplot(meancounts, aes(x = control.mean, y = treated.mean, alpha = 0.2)) +
  geom_point()+
  scale_x_continuous(trans="log2")+
  scale_y_continuous(trans="log2")
  
```
Lets look at genes with a large change between control and treated samples.
```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```
There are some weird NaN results and -Inf results. Lets change our zero values to fix this.

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

Better.

> Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

The purpose of the argument is to eliminate the problem of having zeros come up as "weird"

> Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 

> Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level?

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
sum(up.ind)
sum(down.ind)
```
The answers to 8 and 9 are 250 and 367 respectively.

> Q10. Do you trust these results? Why or why not?

I want to run a significance test before I trust the results.

# 4. DESeq2 analysis

DESeq2 will do all the stats for us.
```{r}
library(DESeq2)
citation("DESeq2")
```

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts, 
                       colData = metadata, 
                       design = ~dex)
dds
```
```{r}
dds <- DESeq(dds)
res <- results(dds)
res
```
```{r}
summary(res)
```

```{r}
head(res)
```
# 6. Data Visualization
```{r}
plot(res$log2FoldChange, res$padj)
```

Well that plot sucked. All the interesting p-values are down below 0.

Lets take the log of p-value

```{r}
plot(res$log2FoldChange, log(res$padj))
```
much better but still weird

lets flip the y axis so its not upside down.

```{r}
plot(res$log2FoldChange, -log(res$padj))
abline( v = c(-2, +2), col = "red")
abline(h=-log(0.05))
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

```
```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

