---
title: "Class 05: Data Visualization with GGPLOT"
format: html
---

#our first ggplot

To use the ggplot 2 package, I first need to have it installed on my computer


To install any package we use the 'install.packages()' command

now can I use it? No, we need to call 'library(ggplot2)'

```{r}
library(ggplot2)
ggplot()
```
```{r}
mpg
```

Our first plot of displ vs hwy
All ggplot() graphs are made in the same way
data+aes+geoms

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  geom_point()

```

I can add more layers 
```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  geom_point()+
  geom_line()
```

Lets make it look nicer

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  geom_point()+
  geom_smooth()
```

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  geom_point()+
  geom_smooth(method =lm, se = FALSE)
```

```{r}
ggplot(cars)
```
```{r}
ggplot(cars) + 
  aes (x=speed, y=dist)
```

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist)+
  geom_point()
```
```{r}
ggplot(cars) + 
  aes(x=speed, y=dist)+
  geom_point()+
  labs(title="Speed and Stopping Distances of Cars", x="speed(MPH)", y="stopping distance(ft)", subtitle ="cars not wars", caption="dataset: 'cars'") +
  geom_smooth(method="lm", se=FALSE)+
  theme_bw()
```




#plot of gene expression data

First read the data from online.

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)

```
```{r}
nrow(genes)
```

```{r}
ncol(genes)
```

```{r}
round(table(genes$State)/nrow(genes)*100,2)
```
```{r}
p <- ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State)+
  geom_point()
p


```
```{r}
p + scale_colour_manual(values=c("red", "orange", "maroon"))+
  labs(title = "gene expression changes upon drug treatment", x = "control (no drug)", y = "drug treatment")
```

