---
title: "Class10"
author: "Joe Kesler"
format: html
---
## 1. Importing candy data
```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
 There are 85 different candy types.
 
 > Q2. How many fruity candy types are in the dataset?
 
```{r}
sum(candy$fruity)
```
 
There are 38 fruity candies.

## 2.What is your favorite candy?

lets look at the variable called winpercent. It shows us who prefers this candy over another randomly chosen candy. Here's an example:
```{r}
candy["Twix", ]$winpercent
```
> Q3.What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy is Haribo Sour Bears. Lets see its winpercent value:
```{r}
candy["Haribo Sour Bears",]$winpercent
```
Its win percent is only 51 :/

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

The people love kit kats

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

The people like tootsie rolls a lot more than I thought.

```{r}
library("skimr")
skim(candy)
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

All the variables that end in "percent" seem to be on a different scale that is continuous. Meanwhile, the non "percent" scales are just between 0 and 1

> Q7. What do you think a zero and one represent for the candy$chocolate column?

0 and 1 represent whether or not the candy fits within the category of the variable. 1 is yes it does fit, and 0 is no it doesn't fit.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) + 
  geom_histogram(bins = 10, col = "brown", fill = "orange")
```

> Q9. Is the distribution of winpercent values symmetrical?

It looks like it is skewed to the right.

> Q10. Is the center of the distribution above or below 50%?

It is below 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate.inds <- as.logical(candy$chocolate)
chocolate.wins <- candy[chocolate.inds,]$winpercent
```
```{r}
fruity.inds <- as.logical(candy$fruity)
fruity.wins <- candy[fruity.inds,]$winpercent
```

```{r}
mean(chocolate.wins)
mean(fruity.wins)
```
The people ON AVERAGE like the chocolate candy better.

> Q12. Is this difference statistically significant?

```{r}
t.test(chocolate.wins, fruity.wins)
```
It is significant because the p-value is so differnt.
People prefer chocolate.

## 3. Overall candy rankings

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```
You can see the 5 least liked above

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent),], n=5)
```
You can see the five most liked above.

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy)+
  aes(winpercent, rownames(candy)) +
  geom_col()
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col() +
  theme(axis.text = element_text(size = 4))
```

Now lets add color.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  theme(axis.text = element_text(size = 4))
```

> Q17. What is the worst ranked chocolate candy?

Sixlets is the worst ranked chocolate candy

> Q18. What is the best ranked fruity candy?

Starburst is the best ranked fruity candy.

## 4. Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

looking at the graph, it seems to be reeses miniatures.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
The top 5 most expensive in the dataset can be seen above. The least popular of these 5 is Nik L Nip

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()+
  theme(axis.text = element_text(size = 3))

```
```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are super anti-correlated. Also bar and pluribus are super anti-correlated

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent are very highly correlated. Bar and chocolate are also extremely correlated.

## 6. PCA

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
The above graph is the PCA for our candies. Lets make this prettier using ggplot though.

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
Lets put in some non overlapping candy names using ggrepel

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 10)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
library(plotly)

ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus all send candies in the PC1 positive direction. this all makes sense because they are all correlated as can be seen in the correlation graph. Also just when you think about from a how candies are packaged point of view, it makes sense.
